```{r}
#Libraries
#install.packages(tidyverse)
#install.packages(plyr)
#install.packages("baseballr")

library(tidyverse)
library(plyr)
library("baseballr")
```
```{r}
plot_sz <- function() {
  x_range = c(-8.5/12, 8.5/12)
  y_range = c(1.5, 3.5)
  segments(x0 = c(8.5/12, -8.5/12, 8.5/12 - 17/36, 17/36 - 8.5/12), y0 = rep(y_range[1], 4), x1 = c(8.5/12, -8.5/12, 8.5/12 - 17/36, 17/36 - 8.5/12), y1 = rep(y_range[2], 4), col = 'black', lwd = 4)
  segments(x0 = rep(x_range[1], 4), y0 = c(1.5, 3.5, 3.5 - 2/3, 3.5 - 4/3), x1 = rep(x_range[2], 4), y1 = c(1.5, 3.5, 3.5 - 2/3, 3.5 - 4/3), col = 'black', lwd = 4)
  segments(x0 = c(8.5/12, -8.5/12, 8.5/12 - 17/36, 17/36 - 8.5/12), y0 = rep(y_range[1], 4), x1 = c(8.5/12, -8.5/12, 8.5/12 - 17/36, 17/36 - 8.5/12), y1 = rep(y_range[2], 4), col = 'white', lwd = 2)
  segments(x0 = rep(x_range[1], 4), y0 = c(1.5, 3.5, 3.5 - 2/3, 3.5 - 4/3), x1 = rep(x_range[2], 4), y1 = c(1.5, 3.5, 3.5 - 2/3, 3.5 - 4/3), col = 'white', lwd = 2)
}
```


```{r}
data <- read.csv("C:/Users/edoug/Downloads/csas_challenge_data.csv")
head(data)
```

**Small samples are not indicative of a player's actual abilities. Therefore, we decided to create a list of players with over 50 plate appearances we could filter future data with.**
```{r}
#Get a list of players over 50 PA's
PA <- data %>% 
  filter(events != "") %>%
  group_by(player_name) %>%
  dplyr::summarize(plate_appearances = n()) %>%
  filter(plate_appearances >= 50) %>%
  arrange(desc(plate_appearances))

over_50 <- as.list(PA$player_name)
```

**In order to perform analysis on bat speed data, we had to create a table with all player stats first. This is so we can eventually check for correlation between basic stats such as batting average and our bat speed data.**
```{r}
#Group all data by player and event to get all events per player (ex: CJ Abrams singles)
stats <- data %>%
  filter(events != "") %>%
  group_by(player_name, events) %>%
  dplyr::summarize(count = n(), .groups = 'drop')

#Use pivot_wider function to make a usable dataframe and calculate rate stats (ex: batting average)
stats_wide <- stats %>%
  pivot_wider(names_from = events, values_from = count) %>%
  replace(is.na(.), 0) %>%
  mutate(hits = single + double + triple + home_run, 
         singles = single, 
         doubles = double, 
         triples = triple, 
         home_runs = home_run,
         walks = walk,
         hbp = hit_by_pitch,
         plate_appearances = hits + walk + hbp + double_play + field_error + field_out + force_out + grounded_into_double_play + sac_fly + strikeout +  fielders_choice +   fielders_choice_out + catcher_interf + sac_bunt + triple_play + strikeout_double_play + sac_fly_double_play, 
         at_bats = plate_appearances - walk - hit_by_pitch - sac_fly - catcher_interf - sac_bunt - sac_fly - sac_fly_double_play,
         avg = round(hits / at_bats, 3),
         obp = round((hits + walks + hbp) / (at_bats + walks + hbp + sac_fly + sac_bunt), 3),
         slg = round((singles + 2 * doubles + 3 * triples + 4 * home_runs) / at_bats, 3),
         ops = obp + slg) %>%
  select(player_name, plate_appearances, at_bats, hits, singles, doubles, triples, home_runs, avg, obp, slg, ops)

stats_wide
```

```{r}
#Get player's average bat speed
avg_bat_speed <- data %>% 
  filter(!is.na(bat_speed)) %>%
  group_by(player_name) %>%
  dplyr::summarize(average_bat_speed = mean(bat_speed)) %>%
  arrange(desc(average_bat_speed))

head(avg_bat_speed)
```

```{r}
#Get player's average bat speed with 2 strikes
two_strikes_bat_speed <- data %>% 
  filter(!is.na(bat_speed)) %>%
  filter(strikes == 2) %>%
  group_by(player_name) %>%
  dplyr::summarize(average_bat_speed = mean(bat_speed)) %>%
  arrange(desc(average_bat_speed))

head(two_strikes_bat_speed)
```

```{r}
bat_speed_df <- avg_bat_speed %>% 
  left_join(two_strikes_bat_speed, by = "player_name")
bat_speed_df
```

**After combining the player stats and bat speed data, we have our table that will be used for the bulk of the analysis. Including players with a small sample of data can create misleading results. We decided to remove any players with less than 50 plate appearances to counteract this trend.**
```{r}
stats_wide <- stats_wide %>%
  filter(player_name %in% over_50)

bat_speed_df <- bat_speed_df %>%
  filter(player_name %in% over_50)

stats_table <- stats_wide %>%
  left_join(bat_speed_df, by = "player_name") %>%
  mutate(avg_bat_speed = average_bat_speed.x,
         avg_two_strike_bat_speed = average_bat_speed.y) %>%
  select(!c(average_bat_speed.x, average_bat_speed.y))

stats_table
```

```{r}
top_25 <- bat_speed_df %>%
  filter(average_bat_speed.x > quantile(bat_speed_df$average_bat_speed.x)[4]) %>%
  select(player_name)
top_25

bot_25 <- bat_speed_df %>%
  filter(average_bat_speed.x < quantile(bat_speed_df$average_bat_speed.x)[2]) %>%
  select(player_name)
bot_25
```

```{r}
zone_data <- data %>%
  select(player_name, plate_x, plate_z, sz_top, sz_bot, zone, description)

plot(zone_data$plate_x, zone_data$plate_z, xlim = c(-5, 5), ylim = c(0, 5), asp = 1, col = zone_data$zone)

#Plot Strike Zone
plot_sz()
```
**Testing data mutation to adjust all pitches based on batter strike zone**
```{r}
zone_data_adj <- zone_data %>%
  mutate(new_plate_z = 1.5 + (plate_z - sz_bot) / (sz_top - sz_bot) * 2)

plot(zone_data_adj$plate_x, zone_data_adj$new_plate_z, xlim = c(-5, 5), ylim = c(0, 5), asp = 1, col = zone_data_adj$zone)

#Plot Strike Zone
plot_sz()
```


```{r}
zone_data_1 <- data %>%
  select(player_name, plate_x, plate_z, zone, sz_top, sz_bot, description, events) %>%
  filter(events %in% c('single', 'double', 'triple', 'home_run', 'field_out', 'field_error', 'fielders_choice', 'fielders_choice_out', 'double_play', 'triple_play', 'other_out', 'force_out', 'strikeout', 'strikeout_double_play', 'grounded_into_double_play')) %>%
  mutate(total_bases = case_when(
    events == "single" ~ 1,
    events == "double" ~ 2,
    events == "triple" ~ 3,
    events == "home_run" ~ 4,
    TRUE ~ 0
  )) %>%
  mutate(group = case_when(
    player_name %in% top_25$player_name ~ 1,
    player_name %in% bot_25$player_name ~ 2,
    TRUE ~ 0
  )) #%>% select(plate_x, plate_y, total_bases, group)

zone_data_1
```
